---
layout: page
title: dplyr and SQL
reading: "<a href='http://cran.r-project.org/web/packages/dplyr/vignettes/databases.html'>dplyr database vignette</a>"
notes: 
output: 
    ioslides_presentation:
        widescreen: true
        smaller: false
slides: true
link: true
---

```{r, echo=FALSE, include=FALSE}
options(width = 110)
library(dplyr)
```


# The why of databases

## Numbers everyone should know

| Task                                | Timing            |
|-------------------------------------|-------------------|
| L1 cache reference                  | 0.5 ns            |
| L2 cache reference                  | 7 ns              |
| Main memory reference               | 100 ns            |
| Read 1 MB sequentially from memory  | 250,000 ns        |
| Disk seek                           | 10,000,000 ns     |
| Read 1 MB sequentially from network | 10,000,000 ns     |
| Read 1 MB sequentially from disk    | 30,000,000 ns     |

<div style="font-size: 10pt">
From: http://surana.wordpress.com/2009/01/01/numbers-everyone-should-know/
</div>


## Implications for bigish data

Lets imagine we have a *10 GB* flat data file and that we want to select certain rows based on a given criteria. This requires a sequential read across the entire data set.


If we can store the file in memory:

* $10~GB \times (250~\mu s / 1~MB) = 0.25$ seconds

If we have to access the file from disk:

* $10~GB \times (30~ms / 1~MB) = 30$ seconds

<br/>

This is just for *reading* data, if we make any modifications (*writing*) things are much worse. 


## Implications for big data

What about a *100 GB* flat data file?

If we can store the file in memory:

* $100~GB \times (250~\mu s / 1~MB) = 2.5$ seconds

If we have to access the file from disk:

* $100~GB \times (30~ms / 1~MB) = 300$ seconds

<br/>

This is actually incredibly optimistic since it assumes that all the data on disk can be read in one continuous read.


## Blocks

<div class="centered">
Cost: Disk << Memory

Speed: Disk <<< Memory
</div>

<br/>

So usually possible to grow our disk storage to accommodate our data. However, memory is usually the limiting resource, and if we can't fit everything into memory?

<br/>

Create *blocks* - group rows based on similar attributes and read in multiple rows at a time. Optimal size will depend on the task and the properties of the disk.



## Linear vs Binary Search

Even with blocks, any kind of subsetting of rows requires a linear search, which requires $\mathcal{O}(N)$ accesses where $N$ is the number of blocks.

<br/>

We can do much better if we are careful about how we structure our data, specifically sorting some or all of the columns. 

* Sorting is expensive, $\mathcal{O}(N \log N)$, but it only needs to be done once. 

* After sorting, we can use a binary search for any subsetting tasks ($\mathcal{O}(\log N)$).

* These sorted columns are known as *indexes*.

* Indexes require additional storage, but usually small enough to be kept in memory while blocks stay on disk.


## Binary Search Example {.smaller}

| Age  | Name         |
|------|--------------|
| 19   | Carol        |
| 20   | Greg         |
| 21   | Alice        |
| 21   | Dave         |
| 22   | Eve          |
| 23   | Bob          |
| 23   | Frank        |

<br/><br/>

Lets search for records for people who are 22 or older.


## and then?

This is just barely scratching the surface,

* Efficiency gains are not just for disk, access is access

* In general, trade off between storage and efficiency

* Reality is a lot more complicated for everything mentioned so far, lots of very smart people have spent a lot of time thinking about and implementing tools

* Different tasks with different requirements require different implementations and have different criteria for optimization



## SQL

Structures Query Language is a special purpose language for interacting with (querying and modifying) these indexed tabular data structures. 

* ANSI Standard but with some dialect divergence

* This functionality maps very closely (but not exactly) with the data manipulation verbs present in dplyr.

* We will see this mapping in more detail in a bit.


# Back to dplyr

## data

```{r eval=FALSE}
library(dplyr)
library(readr)
library(lubridate)
library(stringr)

nyc = read_csv("/data/nyc_parking/NYParkingViolations.csv")

##  |================================================================================| 100% 1713 MB
##   
##  Warning: 654437 parsing failures.
##   row                  col   expected actual
##  2647 Violation Legal Code an integer      T
##  3792 Violation Legal Code an integer      T
##  4001 Violation Legal Code an integer      T
##  4002 Violation Legal Code an integer      T
##  4003 Violation Legal Code an integer      T
##  .... .................... .......... ......
##  .See problems(...) for more details.
```

## {.smaller}

```{r eval=FALSE}
nyc

##  Source: local data frame [9,100,278 x 43]
##  
##     Summons Number Plate ID Registration State Plate Type Issue Date Violation Code
##              (dbl)    (chr)              (chr)      (chr)      (chr)          (int)
##  1      1361929741  FCJ5493                 NY        PAS 12/18/1970             20
##  2      1366962000  63540MC                 NY        COM 02/02/1971             46
##  3      1356906515  GFM1421                 NY        PAS 09/18/1971             40
##  4      1342296217  FYM5117                 NY        SRF 09/18/1971             21
##  5      1342296199  95V6675                 TX        PAS 09/18/1971             21
##  6      1342296187  GCY4187                 NY        SRF 09/18/1971             21
##  7      1337077380  18972BB                 NY        999 10/10/1971             14
##  8      1364523796  WNJ4730                 VA        PAS 04/05/1973             14
##  9      1359914924  68091JZ                 NY        COM 07/22/1973             46
##  10     1355498326  EWV4127                 NY        PAS 08/12/1973             21
##  ..            ...      ...                ...        ...        ...            ...
##  Variables not shown: Vehicle Body Type (chr), Vehicle Make (chr), Issuing Agency (chr), Street
##    Code1 (int), Street Code2 (int), Street Code3 (int), Vehicle Expiration Date (int), Violation
##    Location (chr), Violation Precinct (int), Issuer Precinct (int), Issuer Code (int), Issuer
##    Command (chr), Issuer Squad (chr), Violation Time (chr), Time First Observed (chr), Violation
##    County (chr), Violation In Front Of Or Opposite (chr), House Number (chr), Street Name (chr),
##    Intersecting Street (chr), Date First Observed (int), Law Section (int), Sub Division (chr),
##    Violation Legal Code (int), Days Parking In Effect (chr), From Hours In Effect (chr), To Hours
##    In Effect (chr), Vehicle Color (chr), Unregistered Vehicle? (int), Vehicle Year (int), Meter
##    Number (chr), Feet From Curb (int), Violation Post Code (chr), Violation Description (chr), No
##    Standing or Stopping Violation (chr), Hydrant Violation (chr), Double Parking Violation (chr)
```


## Creating an sqlite database {.smaller}

```{r eval=FALSE}
(db = src_sqlite("/data/nyc_parking/NYParkingViolations.sqlite", create = TRUE))

##  src:  sqlite 3.8.6 [/data/nyc_parking/NYParkingViolations.sqlite]
##  tbls:

nyc_sql = copy_to(db, nyc, temporary = FALSE)
db

##  src:  sqlite 3.8.6 [/data/nyc_parking/NYParkingViolations.sqlite]
##  tbls: nyc, sqlite_stat1

nyc_sql = tbl(db,"nyc")
str(nyc_sql)

##  List of 9
##   $ src      :List of 3
##    ..$ con :Formal class 'SQLiteConnection' [package "RSQLite"] with 5 slots
##    .. .. ..@ Id                 :<externalptr> 
##    .. .. ..@ dbname             : chr "/data/nyc_parking/NYParkingViolations.sqlite"
##    .. .. ..@ loadable.extensions: logi TRUE
##    .. .. ..@ flags              : int 6
##    .. .. ..@ vfs                : chr ""
##    ..$ path: chr "/data/nyc_parking/NYParkingViolations.sqlite"
##    ..$ info:List of 2
##    .. ..$ serverVersion: chr "3.8.6"
##    .. ..$ results      : logi FALSE
##    ..- attr(*, "class")= chr [1:3] "src_sqlite" "src_sql" "src"
##   $ from     :Classes 'ident', 'sql', 'character'  chr "nyc"
## ...
```

## {.smaller}

```{r eval=FALSE}
nyc_sql

##  Source: sqlite 3.8.6 [/data/nyc_parking/NYParkingViolations.sqlite]
##  From: nyc [9,100,278 x 43]
##  
##     Summons Number Plate ID Registration State Plate Type Issue Date Violation Code
##              (dbl)    (chr)              (chr)      (chr)      (chr)          (int)
##  1      1361929741  FCJ5493                 NY        PAS 12/18/1970             20
##  2      1366962000  63540MC                 NY        COM 02/02/1971             46
##  3      1356906515  GFM1421                 NY        PAS 09/18/1971             40
##  4      1342296217  FYM5117                 NY        SRF 09/18/1971             21
##  5      1342296199  95V6675                 TX        PAS 09/18/1971             21
##  6      1342296187  GCY4187                 NY        SRF 09/18/1971             21
##  7      1337077380  18972BB                 NY        999 10/10/1971             14
##  8      1364523796  WNJ4730                 VA        PAS 04/05/1973             14
##  9      1359914924  68091JZ                 NY        COM 07/22/1973             46
##  10     1355498326  EWV4127                 NY        PAS 08/12/1973             21
##  ..            ...      ...                ...        ...        ...            ...
##  Variables not shown: Vehicle Body Type (chr), Vehicle Make (chr), Issuing Agency (chr), Street
##    Code1 (int), Street Code2 (int), Street Code3 (int), Vehicle Expiration Date (int), Violation
##    Location (chr), Violation Precinct (int), Issuer Precinct (int), Issuer Code (int), Issuer
##    Command (chr), Issuer Squad (chr), Violation Time (chr), Time First Observed (chr), Violation
##    County (chr), Violation In Front Of Or Opposite (chr), House Number (chr), Street Name (chr),
##    Intersecting Street (chr), Date First Observed (int), Law Section (int), Sub Division (chr),
##    Violation Legal Code (int), Days Parking In Effect (chr), From Hours In Effect (chr), To Hours
##    In Effect (chr), Vehicle Color (chr), Unregistered Vehicle? (int), Vehicle Year (int), Meter
##    Number (chr), Feet From Curb (int), Violation Post Code (chr), Violation Description (chr), No
##    Standing or Stopping Violation (chr), Hydrant Violation (chr), Double Parking Violation (chr)
```


## Using dplyr with sqlite {.smaller}

```{r, eval=FALSE}
(addr = nyc_sql %>%
        select(`Issue Date`, `Issuing Agency`, `Violation Precinct`, `House Number`, `Street Name`) %>%
        filter(`Violation Precinct` >=1, `Violation Precinct` <= 34)
)

##  Source: sqlite 3.8.6 [/data/nyc_parking/NYParkingViolations.sqlite]
##  From: nyc [3,564,798 x 5]
##  Filter: `Violation Precinct` >= 1, `Violation Precinct` <= 34 
##  
##     Issue Date Issuing Agency Violation Precinct House Number  Street Name
##          (chr)          (chr)              (int)        (chr)        (chr)
##  1  09/18/1971              X                 33         4165     BROADWAY
##  2  07/22/1973              P                 10           48        7 AVE
##  3  09/22/1973              P                 14          205      W 39 ST
##  4  10/30/1973              K                  1           NA SOUTH STREET
##  5  04/15/1977              X                 17          545        1 AVE
##  6  01/01/2000              P                 17          875      3RD AVE
##  7  01/06/2000              P                 20          210      W 64 ST
##  8  01/07/2000              P                 15         1375        6 AVE
##  9  01/07/2000              P                 15         1375        6 AVE
##  10 01/08/2000              P                 19          244      EAST 84
##  ..        ...            ...                ...          ...          ...
```

## SQL Query

```{r, eval=FALSE}
class(addr)

## [1] "tbl_sqlite" "tbl_sql"    "tbl"  

addr$query

##  <Query> SELECT "Issue Date" AS "Issue Date", "Issuing Agency" AS "Issuing Agency", "Violation Precinct" AS "Violation ##  Precinct", "House Number" AS "House Number", "Street Name" AS "Street Name"
##  FROM "nyc"
##  WHERE "Violation Precinct" >= 1.0 AND "Violation Precinct" <= 34.0
##  <SQLiteConnection>
```

## Limitations

```{r, eval=FALSE}
addr %>% mutate(address = paste(`House Number`, `Street Name`))

##  Source: sqlite 3.8.6 [/data/nyc_parking/NYParkingViolations.sqlite]
##  Error in sqliteSendQuery(con, statement, bind.data) : 
##    error in statement: no such function: PASTE

addr %>% summarize(mean = mean(`Violation Precinct`, na.rm=TRUE))

##  Error: na.rm not needed in SQL: NULL are always dropped

addr %>% summarize(mean = mean(`Violation Precinct`))

##  Source: sqlite 3.8.6 [/data/nyc_parking/NYParkingViolations.sqlite]
##  From: <derived table> [?? x 1]
##  
##        mean
##       (dbl)
##  1  16.0982
##  ..     ...
```


## SQL Grouping {.smaller}

```{r, eval=FALSE}
addr %>% group_by(`Issuing Agency`, `Violation Precinct`) %>% summarize(n=n())

##  Source: sqlite 3.8.6 [/data/nyc_parking/NYParkingViolations.sqlite]
##  From: <derived table> [?? x 3]
##  Grouped by: `Issuing Agency` 
##  
##     Issuing Agency Violation Precinct     n
##              (chr)              (int) (int)
##  1               A                  1    13
##  2               A                  7     1
##  3               A                 10    24
##  4               A                 11     1
##  5               A                 14    47
##  6               A                 33    11
##  7               B                 10     1
##  8               B                 25     2
##  9               C                  5    73
##  10              C                 13     7
##  ..            ...                ...   ...
```


## SQL Query

```{r, eval=FALSE}
addr %>% group_by(`Issuing Agency`, `Violation Precinct`) %>% summarize(n=n()) %>% .$query

##  <Query> SELECT "Issuing Agency", "Violation Precinct", "n"
##  FROM (SELECT "Issuing Agency", "Violation Precinct", COUNT() AS "n"
##  FROM "nyc"
##  WHERE "Violation Precinct" >= 1.0 AND "Violation Precinct" <= 34.0
##  GROUP BY "Issuing Agency", "Violation Precinct") AS "zzz4"
##  <SQLiteConnection>
```

## SQL Translation {.smaller}

dplyr has a function, `translate_sql`, that lets you experiment with how R functions are translated to SQL

<div class="columns-2">
```{r, error=TRUE}
translate_sql(x == 1 & (y < 2 | z > 3))
translate_sql(x ^ 2 < 10)
translate_sql(x %% 2 == 10)
translate_sql(paste(x,y))
translate_sql(mean(x))
translate_sql(mean(x, na.rm=TRUE))
```
</div>

<br/>

In general, dplyr knows how to translate basic math, logical, and summary functions from R to SQL.


## (Unfair) Timings {.smaller}

```{r, eval=FALSE}
system.time(
    nyc %>%
    select(`Issue Date`, `Issuing Agency`, `Violation Precinct`, `House Number`, `Street Name`) %>%
    filter(`Violation Precinct` >=1, `Violation Precinct` <= 34) %>%
    group_by(`Issuing Agency`, `Violation Precinct`) %>%
    summarize(n=n())
)

##    user  system elapsed 
##   0.860   0.121   0.981 

system.time(
    nyc_sql %>%
    select(`Issue Date`, `Issuing Agency`, `Violation Precinct`, `House Number`, `Street Name`) %>%
    filter(`Violation Precinct` >=1, `Violation Precinct` <= 34) %>%
    group_by(`Issuing Agency`, `Violation Precinct`) %>%
    summarize(n=n())
)

##    user  system elapsed 
##   0.039   0.001   0.038 
```

`nyc_sql` was 30x times faster than `nyc`, but the former is disk based while the latter is in memory, why this discrepancy?


## Laziness

`dplyr` uses lazy evaluation as much as possible, particularly when working with non-local backends.

* When building a query, we don't want the entire table, often we want just enough to check if our query is working.

* Since we would prefer to run one complex query over many simple queries, laziness allows for verbs to be strung together.

* Therefore, by default `dplyr`
    
    * won't connect and query the database until absolutely necessary (e.g. show output),

    * and unless explicitly told to, will only query a handful of rows to give a sense of what the result will look like

## {.smaller}


```{r eval=FALSE}
nyc_sql %>%
  select(`Issue Date`, `Issuing Agency`, `Violation Precinct`, `House Number`, `Street Name`) %>%
  filter(`Violation Precinct` >=1, `Violation Precinct` <= 34) %>%
  group_by(`Issuing Agency`, `Violation Precinct`) %>%
  summarize(n=n())

##  Source: sqlite 3.8.6 [/data/nyc_parking/NYParkingViolations.sqlite]
##  From: <derived table> [?? x 3]
##  Grouped by: `Issuing Agency` 
##  
##     Issuing Agency Violation Precinct     n
##              (chr)              (int) (int)
##  1               A                  1    13
##  2               A                  7     1
##  3               A                 10    24
##  4               A                 11     1
##  5               A                 14    47
##  6               A                 33    11
##  7               B                 10     1
##  8               B                 25     2
##  9               C                  5    73
##  10              C                 13     7
##  ..            ...                ...   ...
```

## Full query {.smaller}

To force a full query and return a complete `tbl_df` object `dplyr` uses the `collect` function.

```{r, eval=FALSE}
nyc_sql %>%
  select(`Issue Date`, `Issuing Agency`, `Violation Precinct`, `House Number`, `Street Name`) %>%
  filter(`Violation Precinct` >=1, `Violation Precinct` <= 34) %>%
  group_by(`Issuing Agency`, `Violation Precinct`) %>%
  summarize(n=n()) %>%
  collect()

##  Source: local data frame [200 x 3]
##  Groups: Issuing Agency [15]
##  
##     Issuing Agency Violation Precinct     n
##              (chr)              (int) (int)
##  1               A                  1    13
##  2               A                  7     1
##  3               A                 10    24
##  4               A                 11     1
##  5               A                 14    47
##  6               A                 33    11
##  7               B                 10     1
##  8               B                 25     2
##  9               C                  5    73
##  10              C                 13     7
##  ..            ...                ...   ...
```

`compute` and `collapse` also force a full query but have slightly different behavior and return types. 


## Creating Indexes


```{r, eval=FALSE}
(db_index = src_sqlite("/data/nyc_parking/NYParkingViolations_index.sqlite", create = TRUE))

## src:  sqlite 3.8.6 [/data/nyc_parking/NYParkingViolations_index.sqlite]
## tbls:

nyc_index = copy_to(db_index, nyc, temporary = FALSE,
                    index = list(`Violation Precinct`))
```


##

The indexed database takes up more disk space:

```shell
cr173@gort [~]$ ls -lh /data/nyc_parking/*.sqlite

-rw-r--r-- 1 cr173 visitor 1.8G Mar 30 11:56 NYParkingViolations_index.sqlite
-rw-r--r-- 1 cr173 visitor 1.7G Mar 30 11:28 NYParkingViolations.sqlite
```

## Timings for filtering precincts

```{r, eval=FALSE}
system.time(nyc_sql %>% filter(`Violation Precinct` <= 34, `Violation Precinct` >= 1) %>% collect())

##    user  system elapsed 
##  30.630   3.279  34.009 

system.time(nyc_index %>% filter(`Violation Precinct` <= 34, `Violation Precinct` >= 1) %>% collect())

##    user  system elapsed 
##  32.207   3.197  35.395 
```

## Timings for grouping

```{r, eval=FALSE}
system.time(nyc_sql %>% group_by(`Violation Precinct`) %>% summarize(n=n()) %>% collect())

##     user  system elapsed 
##   21.184   2.829  24.007 

system.time(nyc_index %>% group_by(`Violation Precinct`) %>% summarize(n=n()) %>% collect())

##   user  system elapsed 
##  1.288   0.102   1.389 
```


# Acknowledgments

## Acknowledgments

Above materials are derived in part from the following sources:

* [dplyr - Introduction Vignette](http://cran.r-project.org/web/packages/dplyr/vignettes/introduction.html)
* [dplyr - Databases Vignette](http://cran.r-project.org/web/packages/dplyr/vignettes/databases.html)