---
layout: page
title: ggplot2
reading: "<a href='http://docs.ggplot2.org/current/'>ggplot2 documentation</a> <br/> <a href='https://www.rstudio.com/wp-content/uploads/2015/12/ggplot2-cheatsheet-2.0.pdf'>Visualization Cheatsheet</a>"
notes: "Screencast - <a href='https://youtu.be/d01X6OcbGkQ'>ggplot2 exercises</a>"
output: 
    ioslides_presentation:
        widescreen: true
slides: true
link: true
---

```{r, echo=FALSE, results='hide'}
options(width = 100)

suppressMessages(library(dplyr))
suppressMessages(library(ggplot2))

knitr::opts_chunk$set(fig.align="center", warning=FALSE, message=FALSE)
```

# ggplot2

## ggplot2

- ggplot2 is a plotting system for R, based on the grammar of graphics, which tries to take the good parts of base and lattice graphics and none of the bad parts. 

- It takes care of many of the fiddly details that make plotting a hassle (like drawing legends) as well as providing a powerful model of graphics that makes it easy to produce complex multi-layered graphics.

<br>

Source: http://ggplot2.org/

## Why ggplot2?

- **Statistical:** Visualize $Y \sim X | Z$ with ease

- **Technical:** Human readable syntax

- **Bonus:** Aesthetically pleasing


## New data - movies {.smaller}

The data set is comprised of 651 randomly sampled movies produced and 
released before 2016. Data come from [IMDB](http://www.imdb.com/) and 
[Rotten Tomatoes](http://www.rottentomatoes.com/). Codebook is available [here](http://www.stat.duke.edu/~cr173/Sta523_Fa16/data/movies/movies.html).


```{r}
movies = read.csv("http://www.stat.duke.edu/~cr173/Sta523_Fa16/data/movies/movies.csv", 
                  stringsAsFactors = FALSE) %>% tbl_df()
movies
```



## With ggplot2 {.smaller}

```{r fig.height = 4}
ggplot(data = movies, aes(x = critics_score, y = audience_score, color = title_type)) +
  geom_point(alpha=0.5) +
  geom_smooth(method = "lm", se = FALSE)
```

## With base R {.smaller}

```{r eval=FALSE}
plot(y = movies$audience_score, x = movies$critics_score, 
     col = adjustcolor(as.integer(factor(movies$title_type)),alpha=0.5), pch=16)

doc = movies[movies$title_type == "Documentary", ]
ff = movies[movies$title_type == "Feature Film", ]
tv = movies[movies$title_type == "TV Movie", ]

m_doc = lm(audience_score ~ critics_score, data = doc)
m_ff =  lm(audience_score ~ critics_score, data = ff)
m_tv =  lm(audience_score ~ critics_score, data = tv)

abline(m_doc, col = 1, lwd=2)
abline(m_ff, col = 2, lwd=2)
abline(m_tv, col = 3, lwd=2)

legend("topleft", levels(factor(movies$title_type)), 
       col = c(1,2,3), lty = 1)
```

## With base R {.smaller}

```{r echo=FALSE}
plot(y = movies$audience_score, x = movies$critics_score, 
     col = adjustcolor(as.integer(factor(movies$title_type)),alpha=0.5), pch=16)

doc = movies[movies$title_type == "Documentary", ]
ff = movies[movies$title_type == "Feature Film", ]
tv = movies[movies$title_type == "TV Movie", ]

m_doc = lm(audience_score ~ critics_score, data = doc)
m_ff =  lm(audience_score ~ critics_score, data = ff)
m_tv =  lm(audience_score ~ critics_score, data = tv)

abline(m_doc, col = 1, lwd=2)
abline(m_ff, col = 2, lwd=2)
abline(m_tv, col = 3, lwd=2)

legend("topleft", levels(factor(movies$title_type)), 
       col = c(1,2,3), lty = 1)
```



# Grammar of Graphics

## The Grammar of Graphics

- Visualisation concept created by Wilkinson (1999)
    - to define the basic elements of a statistical graphic 

- Adapted for R by Wickham (2009) who created the `ggplot2` package
    - consistent and compact syntax to describe statistical graphics
    - highly modular as it breaks up graphs into semantic components 

- It is not meant as a guide to which graph to use and how to best convey your data (more on that next week).

<br>

Source: https://rpubs.com/timwinke/ggplot2workshop

## The Grammar of Graphics - Terminology

A statistical graphic is a...

- mapping of **data**
- to **aesthetic attributes** (color, size, xy-position)
- using **geometric objects** (points, lines, bars)
- with data being **statistically transformed** (summarised, log-transformed)
- and mapped onto a specific **facet** and **coordinate system**

## Audience score vs. critics score {.smaller}

```{r echo=FALSE, fig.height=3, fig.width=5, warning=FALSE}
ggplot(data = movies, aes(x = audience_score, y = critics_score)) +
  geom_point()
```

- Which data is used as an input?
- What geometric objects are chosen for visualization?
- What variables are mapped onto which attributes?
- What type of scales are used to map data to aesthetics?
- Are the variables statistically transformed before plotting?

## Audience score vs. critics score - code {.smaller}

```{r fig.height=3, fig.width=5}
ggplot(data = movies, aes(x = audience_score, y = critics_score)) +
  geom_point()
```

## Altering features {.smaller}

```{r echo=FALSE, fig.height=3, fig.width=5, warning=FALSE}
ggplot(data = movies, aes(x = audience_score, y = critics_score)) +
  geom_point(alpha = 0.5, color = "blue")
```

- How did the plot change?
- Are these changes based on data (i.e. can be mapped to variables in the dataset) or
are the changes based on stylistic choices for the geometric objects??

## Altering features - code

```{r fig.height=3, fig.width=5}
ggplot(data = movies, aes(x = audience_score, y = critics_score)) +
  geom_point(alpha = 0.5, color = "blue")
```

## Faceting {.smaller}

```{r echo=FALSE, fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = audience_score, y = critics_score, color = genre)) +
  geom_point(alpha = 0.5) +
  facet_grid(. ~ title_type)
```

- How did the plot change?
- Are these changes based on data (i.e. can be mapped to variables in the dataset) or
are the changes based on stylistic choices for the geometric objects?

## Faceting - code {.smaller}

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = audience_score, y = critics_score, color = genre)) +
  geom_point(alpha = 0.5) +
  facet_grid(. ~ title_type)
```

## More faceting {.smaller}

```{r fig.height=4, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = audience_score, y = critics_score, color = genre)) +
  geom_point(alpha = 0.5) +
  facet_grid(audience_rating ~ title_type)
```

## Even More faceting {.smaller}

```{r fig.height=4, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = audience_score, y = critics_score, color = title_type)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~genre)
```

## Anatomy of a ggplot

```
ggplot(data = [dataframe], aes(x = [var_x], y = [var_y], 
       color = [var_for_color], fill = [var_for_fill], 
       shape = [var_for_shape])) +
    geom_[some_geom]([geom_arguments]) +
    ... # other geometries
    scale_[some_axis]_[some_scale]() +
    facet_[some_facet]([formula]) +
    ... # other options
```

# Various plots

## Histograms

```{r fig.height=3, fig.width=7, warning=FALSE}
ggplot(data = movies, aes(x = audience_score)) +
  geom_histogram(binwidth = 5)
```

## Boxplots

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(y = audience_score, x = genre)) +
  geom_boxplot()
```

## Boxplots - axis formatting

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(y = audience_score, x = genre)) +
  geom_boxplot() +
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

## Density plots - border color

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = runtime, color = audience_rating)) +
  geom_density() 
```

## Density plots - fill color

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = runtime, fill = audience_rating)) +
  geom_density() 
```

## Density plots - fill color, with alpha

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = runtime, fill = audience_rating)) +
  geom_density(alpha = 0.5) 
```

## Scatter plots

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = imdb_rating, y = audience_score)) +
  geom_point(alpha = 0.5) 
```

## Smoothing

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = imdb_rating, y = audience_score)) +
  geom_point(alpha = 0.5) +
  geom_smooth()
```

## Smoothing - lm

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = imdb_rating, y = audience_score)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm")
```

## Barplots

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = genre)) +
  geom_bar() +
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

## Segmented barplots

```{r fig.height=3, fig.width=9, warning=FALSE}
ggplot(data = movies, aes(x = genre, fill = audience_rating)) +
  geom_bar() +
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

## Segmented barplots - proportions

```{r fig.height=3, fig.width=9, warning=FALSE}
ggplot(data = movies, aes(x = genre, fill = audience_rating)) +
  geom_bar(position = "fill") +
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

## Dodged barplots

```{r fig.height=3, fig.width=9, warning=FALSE}
ggplot(data = movies, aes(x = genre, fill = audience_rating)) +
  geom_bar(position = "dodge") +
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

## Not on team `theme_grey()`?

```{r fig.height=3, fig.width=9, warning=FALSE}
ggplot(data = movies, aes(x = genre, fill = audience_rating)) +
  geom_bar(position = "dodge") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## More `ggplot2` resources

- Visit http://docs.ggplot2.org/current/ for documentation on the current version
of the `ggplot2` package. It's full of examples!

- Refer to the `ggplot2` cheatsheet: http://www.rstudio.com/wp-content/uploads/2015/12/ggplot2-cheatsheet-2.0.pdf

- Themes vignette: http://docs.ggplot2.org/dev/vignettes/themes.html

## Exercise 1

Recreate the following plot. Hint: Add a `labs()` layer.

```{r echo=FALSE, fig.height=4}
ggplot(data = movies, aes(x = imdb_num_votes, 
                          y = imdb_rating, 
                          color = audience_rating)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~ mpaa_rating) +
  labs(x = "IMDB number of votes", 
       y = "IMDB rating", 
       title = "IMDB and RT scores, by MPAA rating",
       color = "Audience rating") +
  theme_bw()
```

## Exercise 2

Recreate the following plot. Hint: the black lines are a linear model (lm) fit to *all* of the movies within a rating category and the grey line is the 0-1 line (intercept 0, slope 1).

```{r echo=FALSE, fig.height=4}
ggplot(movies, aes(x=audience_score, y=critics_score)) + 
  geom_point(aes(color=best_pic_nom)) + 
  facet_wrap(~mpaa_rating) + 
  geom_smooth(method = "lm",color="black",se=FALSE,alpha=0.5,size=0.5,fullrange=TRUE,show.legend=FALSE)+
  geom_abline(intercept=0,slope=1,color="grey",alpha=0.5)+
  theme_bw()+
  xlim(0,100)+
  ylim(0,100)
```


# Acknowledgments

## Acknowledgments

Above materials are derived in part from the following sources:

* Mine Cetinkaya-Rundel's [DataFest 2016 Visualization Workshop](https://github.com/mine-cetinkaya-rundel/df2016_workshops/tree/master/viz_ggplot2_shiny)

* [RStudio Data Visualization Cheat Sheet](http://www.rstudio.com/wp-content/uploads/2015/12/ggplot2-cheatsheet-2.0.pdf)

* Package Documentation - [ggplot2](http://docs.ggplot2.org/current/), [shiny](http://shiny.rstudio.com/)

* Tim Winkle's [ggplot2 Workshop](https://rpubs.com/timwinke/ggplot2workshop)
