---
layout: page
title: Web APIs & regex
reading: "<a href='http://www.cheatography.com/davechild/cheat-sheets/regular-expressions/'>RegEx Cheat Sheet</a> and <a href='https://cran.r-project.org/web/packages/stringr/vignettes/stringr.html'>stringr</a>"
notes: "<a href='http://regexr.com/'>Testing RegExs</a> <br/> Screencast - <a href='https://youtu.be/gUCRtnoBxik'>regex</a>, <a href='https://youtu.be/vxrOtEwbFoQ'>Dennys</a>"

output: 
    ioslides_presentation:
        widescreen: true
slides: true
link: true
---

```{r echo=FALSE}
suppressMessages(library(stringr))
```

# stringr

## stringr {.smaller}

stringr is a string handling package written by Hadley Wickham that is designed to improve / simplify string handling in R. Most are wrappers for base R functions. 

|                                           |                                     |
|-------------------------------------------|-------------------------------------|
|`str_detect(string, pattern)`              | Detect the presence or absence of a pattern in a string. |
|`str_locate(string, pattern)`              | Locate the first position of a pattern and return a matrix with start and end. |
|`str_extract(string, pattern)`             | Extracts text corresponding to the first match. |
|`str_match(string, pattern)`               | Extracts capture groups formed by `()` from the first match. |
|`str_split(string, pattern)`               | Splits string into pieces and returns a list of character vectors. |
|`str_replace(string, pattern, replacement)`| Replaces the first matched pattern and returns a character vector. |

<br/>

Many of these functions have variants with an `_all` suffix which will match more than one occurrence of the pattern in a given string.


# Regular Expressions

## Simple Pattern Detection

```{r}
text = c("The","quick","brown","fox","jumps","over","the","lazy","dog")
```

```{r}
str_detect(text,"quick")
str_detect(text,"o")
str_detect(text,"row")
```

## Escape Characters {.smaller}

An escape character is a character which results in an alternative interpretation of the following character(s). These vary from language to language but most often `\` is the escape character and it is followed by a single character.

Some common examples:

|    |                  |
|----|------------------|
|`\'`| single quote     |
|`\"`| double quote     |
|`\\`| backslash        |
|`\n`| new line         |
|`\r`| carriage return  |
|`\t`| tab              |
|`\b`| backspace        |
|`\f`| form feed        |


## Metacharacters {.smaller}

The power of regular expressions comes from the ability to use a number of special metacharacters that modify how pattern matching is performed.

```regex
. ^ $ * + ? { } [ ] \ | ( )
```

Because of their special properties they cannot be matched directly, if you need to match one of them you need to escape it first (precede it by `\`). One problem is that regex escapes live on top of regular string escapes, so the following causes an error

```r
str_detect("abc[def","\[")
```
We actually need to prefix `[` with `\\` because `\` is also an escape for normal character strings.

```{r error=TRUE}
str_detect("abc[def","\\[")
```


## Pop quiz {.build}

How do you detect if a string contains a `\` character?

```{r}
cat("abc\\def\n")
str_detect("abc\\def","\\\\")
```

## Anchors

Sometimes we want to specify that our pattern ours at a particular location, we can do this using anchors.

|      | |
|------|-|
| `^` or `\A` | Start of string   |
| `$` or `\Z` | End of string     |
| `\b`        | Word boundary     |  
| `\B`        | Not word boundary |

## Anchor Examples {.smaller}

```{r}
text = "the quick brown fox jumps over the lazy dog"
```
<div class="columns-2">
```{r}
str_replace(text,"^the","---")
str_replace(text,"^dog","---")
str_replace(text,"the$","---")
str_replace(text,"dog$","---")
str_replace(text,"\\Athe","---")
str_replace(text,"dog\\Z","---")
str_replace(text,"\\brow\\b","---")
str_replace(text,"\\Brow\\B","---")
```
</div>


## Character Classes

When we want to match whole classes of characters at a time there are a number of convenience patterns built in,

|      | |
|------|-|
| `\s` | White space |
| `\S` | Not white space |
| `\d` | Digit (0-9)|
| `\D` | Not digit |
| `\w` | Word (A-Z, a-z, 0-9, or _) |
| `\W` | Not word |

## Ranges

We can also specify our own character groups through the construction of ranges

|          | |
|----------|-|
| `.`      | Any character except new line (\n) |
| `[abc]`  | Range (a or b or c) |
| `[^abc]` | Not (a or b or c) |
| `[a-q]`  | Lower case letter from a to q |
| `[A-Q]`  | Upper case letter from A to Q |
| `[0-7]`  | Digit from 0 to 7 |


## Quantifiers

Attached to literals, character classes, ranges or groups to match repeats.

|         | |
|---------|-|
| `*`     | Match 0 or more |
| `+`     | Match 1 or more |
| `?`     | Match 0 or 1 |
| `{3}`   | Match Exactly 3 |
| `{3,}`  | Match 3 or more |
| `{3,5}` | Match 3, 4 or 5 |

## Greedy vs ungreedy matching 

Add a ? to a quantifier to make it ungreedy.

```{r}
text = "<div class='main'> <div> <a href='here.pdf'>Here!</a> </div> </div>"
```

```{r}
str_extract(text, "<div>.*</div>")
str_extract(text, "<div>.*?</div>")
```

## Groups

Group together parts of a regular expression for modification or capture.

|           | |
|-----------|-|
| `(a|b)`   | match literal a or b, group either
| `a(bc)?`  | match literal a or abc, group bc or ""
| `(?:abc)` | Non-c­apt­uring group
| `(abc)def(hig)| match abcdefhig, group abc and hig

## Group Examples {.smaller}

```{r}
text = c("apple", "219 733 8965", "329-293-8753", "Work: 579-499-7527; Home: 543.355.3679")
phone = "([2-9][0-9]{2})[- .]([0-9]{3})[- .]([0-9]{4})"
```

```{r}
str_extract(text, phone)
str_extract_all(text, phone)
```


## Matching {.smaller}

```{r}
str_match(text[2:4], phone)
str_match_all(text[2:4], phone)
```

## How not to use a RegEx {.build}

Validating an email address:

```
(?:[a-z0-9!#$%&'*+/=?^_`{|}~-]+(?:\.[a-z0-9!#$%&'*+/=?^_`{|}~-]+)*|"(?:[\x01-\x08\x0b\x0c\x0e-\x1f\x21\x23-\x5b\x5d-\x7f]|\\[\x01-\x09\x0b\x0c\x0e-\x7f])*")@(?:(?:[a-z0-9](?:[a-z0-9-]*[a-z0-9])?\.)+[a-z0-9](?:[a-z0-9-]*[a-z0-9])?|\[(?:(?:25[0-5]|2[0-4][0-9]|[01]?[0-9][0-9]?)\.){3}(?:25[0-5]|2[0-4][0-9]|[01]?[0-9][0-9]?|[a-z0-9-]*[a-z0-9]:(?:[\x01-\x08\x0b\x0c\x0e-\x1f\x21-\x5a\x53-\x7f]|\\[\x01-\x09\x0b\x0c\x0e-\x7f])+)\])
```



## Exercise 1

For the following vector of randomly generated names, write a regular expression that,

* detects if the person's first name starts with a vowel (a,e,i,o,u)

* detects if the person's last name starts with a vowel

* detects if either the person's first or last name start with a vowel

* detects if neither the person's first nor last name start with a vowel


```{r echo=FALSE}
library(randomNames)
set.seed(124)
dput(randomNames(20, name.order="first.last", name.sep=" "))
```


## Exercise 2 {.smaller}

Write one or more regular expressions to extract the data contained in the xml file below:

```xml
<breakfast_menu>
  <food>
    <name>Belgian Waffles</name>
    <price>$5.95</price>
    <calories>650</calories>
    <addon>
        <name>Strawberries</name>
        <price>$2.00</price>
        <calories>250</calories>
    </addon>
  </food>
  <food>
    <name>French Toast</name>
    <price>$4.50</price>
    <calories>600</calories>
    <addon>
        <name>Strawberries</name>
        <price>$2.00</price>
        <calories>250</calories>
    </addon>
  </food>
</breakfast_menu>
```

# Acknowledgments

## Acknowledgments

* Hadley Wickham - [stringr vigneete](https://cran.r-project.org/web/packages/stringr/vignettes/stringr.html)

* David Child - [RegEx Cheat Sheet](http://www.cheatography.com/davechild/cheat-sheets/regular-expressions/)

* [Regular-Expression.info](http://www.regular-expressions.info/)